---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# basepenguins

<!-- badges: start -->

[![R-CMD-check](https://github.com/EllaKaye/basepenguins/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/EllaKaye/basepenguins/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/EllaKaye/basepenguins/graph/badge.svg)](https://app.codecov.io/gh/EllaKaye/basepenguins)

<!-- badges: end -->

Convert scripts that use the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/index.html) library to use the versions of the `penguins` and `penguins_raw` datasets that are available in R >= 4.5.0.

The Palmer Penguins data has become [very popular](https://apreshill.github.io/palmerpenguins-useR-2022/) in the R community, especially in educational contexts, and as an [alternative to the `iris` dataset](https://journal.r-project.org/articles/RJ-2022-020/). Including the datasets in R makes them more widely available, and easier to get started with, especially for new R users, and for use in teaching.

The version of `penguins` in **datasets** (R >= 4.5.0) has some shorter variable names than the **palmerpenguins** equivalent (e.g. `bill_len` instead of `bill_length_mm`). 
It does mean, however, that for those wanting to use R's version of `penguins`, it isn't simply a case of removing the call to `library(palmerpenguins)` or `data("penguins", package = "palmerpenguins")` and the script still running. The **basepenguins** package takes care of converting files to remove the use of **palmerpenguins** and makes the necessary conversions to variable names, ensuring that the resulting scripts still run.

## Installation

The stable version of **basepenguins** is available on CRAn:

```r
install.packages("basepenguins")
```

Or get the development version from R-universe:

``` r
install.packages("basepenguins", repos = "https://ellakaye.r-universe.dev")
```

## Converting a file

If a file is 'convertible', i.e. it contains `library(palmerpenguins)` or `data("penguins", package = "palmerpenguins")` (with any style of quotes), and has one of a specified set of extensions (by default `"R"`, `"r"`, `"qmd"`, `"rmd"`, `"Rmd"`), then converting it will do the following:

- Replace `library(palmerpenguins)` (or same with `palmerpenguins` in quotes) with the empty string`""`
- Replace `data("penguins", package = "palmerpenguins")` (with any style of quotes) with `data("penguins", package = "datasets")` 
-   Replace variable names:
    -   `bill_length_mm` -\> `bill_len`
    -   `bill_depth_mm` -\> `bill_dep`
    -   `flipper_length_mm` -\> `flipper_len`
    -   `body_mass_g` -\> `body_mass`
-   Replace `ends_with("_mm")` with `starts_with("flipper_"), starts_with("bill_")`

Here, we simply show the 'before-and-after' when converting a single file. For a more extensive guide to using the package, see the [Get Started](https://ellakaye.github.io/basepenguins/articles/basepenguins.html) vignette, `vignette("basepenguins")`.

```{r, include = FALSE}
# Set up a temp directory and set it as the root directory for all chunks
temp_dir <- tempdir()
knitr::opts_knit$set(root.dir = temp_dir)
```

```{r}
library(basepenguins)
```

Get and see an example file provided by the package (adapted from a section of the **palmerpenguins** [Get Started](https://allisonhorst.github.io/palmerpenguins/articles/intro.html) vignette):

```{r}
penguin_file <- example_files("penguins.R")
cat(readLines(penguin_file), sep = "\n")
```

Then, convert the file (saving it to a new file) and see the new script:

```{r}
convert_files(penguin_file, "penguins_converted.R")
cat(readLines("penguins_converted.R"), sep = "\n")
```

## Converting multiple files or a directory

There are four functions in **basepenguins** to convert mulitple files. 
In each case, the default `extensions` (i.e. file types to convert) are `"R"`, `"r"`, `"qmd"`, `"rmd"`, `"Rmd"`.
If `input` contains non-convertible files (i.e. without the specified extensions or without a call to `library(palmerpenguins)`),
they will be copied unmodified to their new `output` location (in `convert_files()` and `convert_dir()` 
or left untouched by `convert_files_inplace()` and `convert_dir_inplace()`).

|  |  |
|----|----|
| `convert_files(input, output, extensions)` | convert an `input` vector of files to new `output` locations |
| `convert_files_inplace(input, extensions)` | convert a vector of files by overwriting them |
| `convert_dir(input, output, extensions)` | convert all files in `input` directory  into a new `output` directory (preserving nesting structure)  |
| `convert_dir_inplace(input, extensions)` | convert all files in a directory by overwriting them |

