---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# basepenguins

<!-- badges: start -->
[![R-CMD-check](https://github.com/EllaKaye/basepenguins/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/EllaKaye/basepenguins/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/EllaKaye/basepenguins/graph/badge.svg)](https://app.codecov.io/gh/EllaKaye/basepenguins)
<!-- badges: end -->

Convert scripts that use the **[palmerpenguins](https://allisonhorst.github.io/palmerpenguins/index.html)** library to use the versions of the `penguins` and `penguins_raw` datasets that are now in R-devel, so coming in R 4.5.

The Palmer Penguins data has become [very popular](https://apreshill.github.io/palmerpenguins-useR-2022/) in the R community, 
especially in educational contexts, and as an [alternative to the `iris` dataset](https://journal.r-project.org/articles/RJ-2022-020/).
Including the datasets in R makes them more widely available, and easier to get started with,
especially for new R users, and for use in teaching.

The version of `penguins` in R's **datasets** package has some shorter variable names
than the **palmerpenguins** equivalent. The differences are:

- `bill_dep` instead of `bill_depth_mm`
- `bill_len` instead of `bill_length_mm`
- `flipper_len` instead of `flipper_length_mm`
- `body_mass` instead of `body_mass_g`

These changes were made for more compact code and data display.
(e.g. printing `penguins` now shows all columns in under 80 characters).

It does mean, however, that for those wanting to use R's version of `penguins`,
it isn't simply a case of removing the call to `library(palmerpenguins)` and the script still running.
The **basepenguins** package takes care of converting files to remove the call to `library(palmerpenguins)` 
and makes the necessary conversions to variable names, ensuring that the resulting scripts still run. 

## Installation

You can install the development version of basepenguins from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("EllaKaye/basepenguins")
```

## Converting a file

If a file contains `library(palmerpenguins)`, and is 'convertible', i.e. 
has one of a specified set of extensions (by default .R, .r, .qmd, .rmd, .Rmd), 
then converting it will do the following:

- Replace `library(palmerpenguins)` with `""` 
- Replace variable names:
  - `bill_length_mm` -> `bill_len`
  - `bill_depth_mm` -> `bill_dep`
  - `flipper_length_mm` -> `flipper_len`
  - `body_mass_g` -> `body_mass`
- Replace `ends_with("_mm")` with `starts_with("flipper_"), starts_with(bill_)`

Here, we simply show the 'before-and-after' when converting a single file.

For a more extensive guide to using the pacakge, 
see the [Get Started](https://ellakaye.github.io/basepenguins/articles/basepenguins.html) vignette, `vignette("basepenguins")`.

```{r, include = FALSE}
# Set up a temp directory and set it as the root directory for all chunks
temp_dir <- tempdir()
knitr::opts_knit$set(root.dir = temp_dir)
```

```{r}
library(basepenguins)
```

Get and see an example file provided by the package:

```{r}
penguin_file <- example_files("penguins.R")
cat(readLines(penguin_file), sep = "\n")
```

Then, convert the file (saving it to a new file) and see the new script:

```{r}
convert_files(penguin_file, "penguins_converted.R")
cat(readLines("penguins_converted.R"), sep = "\n")
```