---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# basepenguins

<!-- badges: start -->
[![R-CMD-check](https://github.com/EllaKaye/basepenguins/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/EllaKaye/basepenguins/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/EllaKaye/basepenguins/graph/badge.svg)](https://app.codecov.io/gh/EllaKaye/basepenguins)
<!-- badges: end -->

Convert scripts that use the **[palmerpenguins](https://allisonhorst.github.io/palmerpenguins/index.html)** library to use the versions of the `penguins` and `penguins_raw` datasets that are now in R-devel, so coming in R 4.5.

The Palmer Penguins data has become [very popular](https://apreshill.github.io/palmerpenguins-useR-2022/) in the R community, 
especially in educational contexts, and as an [alternative to the `iris` dataset](https://journal.r-project.org/articles/RJ-2022-020/).
Including the datasets in R makes them more widely available, and easier to get started with,
especially for new R users, and for use in teaching.

The version of `penguins` in R's **datasets** package has some shorter variable names
than the **palmerpenguins** equivalent. The differences are:

- `bill_dep` instead of `bill_depth_mm`
- `bill_len` instead of `bill_length_mm`
- `flipper_len` instead of `flipper_length_mm`
- `body_mass` instead of `body_mass_g`

These changes were made for more compact code and data display.
(e.g. printing `penguins` now shows all columns in under 80 characters).

It does mean, however, that for those wanting to use R's version of `penguins`,
it isn't simply a case of removing the call to `library(palmerpenguins)` and the script still running.
The **basepenguins** package takes care of converting files to remove the call to `library(palmerpenguins)` 
and makes the necessary conversions to variable names, ensuring that the resulting scripts still run. 

## Installation

You can install the development version of basepenguins from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("EllaKaye/basepenguins")
```

## Usage

This is a quick tour. 
For a more extensive guide, see the [Get Started](https://ellakaye.github.io/basepenguins/articles/basepenguins.html) vignette, `vignette("basepenguins")`.

```{r}
library(basepenguins)
```

## Converting files

```{r}
# get absolute paths of example files
input <- example_files(full.names = TRUE)

# See one of the input files
cat(readLines(input[2]), sep = "\n") 
```

```{r}
# generate output file paths (by default prefix "_new" to input filenames)
output <- extend_names(input) 
```

```{r}
# convert the files
result <- convert_files(input, output)
```

```{r}
# See which files have changed
result
```

```{r}
# See the changes to the input file we saw above
cat(readLines(output[2]), sep = "\n") 
```

Conversions can also be run in-place, which will overwrite files.
Here's how do to that, limited to R scripts, though we don't execute the chunk below:

```{r eval = FALSE}
convert_files_inplace(input, extensions = "R")
```

## Coverting a directory

Whilst `convert_files()` allows the conversion of specified files,
`convert_dir()` will convert all the files (with the stated extensions) in a given directory (and its subdirectories).

```{r}
example_dir <- example_dir()
output_dir <- tempdir()
result <- convert_dir(example_dir, output_dir)

result
```

There is also a version to modify files in place (not executed here):


```{r eval = FALSE}
convert_dir_inplace(example_dir)
```