---
title: "basepenguins"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basepenguins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(basepenguins)
```

**TODO: finish writing this:**

- full demonstration of package functionality
- justification of the `ends_with("_mm")` substitution
- explanation of the package messages

This package provides functions that will take a vector of files (`convert_files()`) or a directory (`convert_dir()`), 
and for all files with specified extensions (by default .R/.r/.qmd/.rmd/.Rmd ), 
remove any calls to `library(palmerpenguins)` 
and replace long palmerpenguins variable names with the shorter R equivalents.
It also deals with `ends_with("_mm")` (a selector used in the [palmerpenguins Get Started vignette](https://allisonhorst.github.io/palmerpenguins/articles/intro.html)).

The versions of `penguins_raw` in **palmerpenguins** and **datasets** are identical, 
expect that in the former it'll have class `tbl_df` if the [tibble](https://tibble.tidyverse.org) package is installed.
No specific changes are made to `penguins_raw` in **basepenguins**, 
but by removing the call to `library(palmerpenguins)`, the **datasets** version will be used in any scripts,
which is always a `data.frame` (never a `tbl_df`).

Note that the **palmerpenguins** package provides features that are not in R,
such as vignettes and articles on the [package website](https://allisonhorst.github.io/palmerpenguins/index.html). The package also contains the data in two csv files and provides a [function](https://allisonhorst.github.io/palmerpenguins/reference/path_to_file.html) to access them. And, of course, Allison Horst's wonderful [penguins artwork](https://allisonhorst.github.io/palmerpenguins/articles/art.html)!
